Revisiting Multi-Domain Machine Translation

### 2020-04-10 2020/coling2020/multidomain.tex
clarify concepts, add references, start reshaping the first part

Owing to the great variability of language data, this ideal situation is rarely met in practice, warranting the study of an alternative scenario, where the test distribution $\mathcal{D'}$ differs from $\mathcal{D}$. In this setting, \emph{domain adaptation} (DA) methods are required. DA has a long history in Machine Learning in general (eg.\ \cite{Shimodaira00improving,Ben-David09atheory}) and in NLP in particular (eg.\ \cite{Daume06domain,Blitzer07domain,Jiang07instance}), and various techniques exist to handle both the situations where a (small) training sample drawn from $\mathcal{D'}$ is available in training, or where only samples of source-side (or target-side) sentences are available (see \cite{Foster07mixture,Bertoldi09domain,Axelrod11domain} for early proposals for statistical MT, or \cite{Chu18asurvey} for a recent survey of DA in the context of Neural MT).
% In a nutshell, DA attempts to compensate for the distribution mismatch either by biasing the training sample to match ${\mathcal{D'}$ (eg.\ using resampling or instance-weighting strategies), or by adapting the

A seemingly related problem is \emph{multi-domain (MD) machine translation} \cite{Hassan17neural,Farajian17multidomain,Kobus17domaincontrol,Zeng18multidomain,Pham19generic} where a single system has to be trained  and tested with data from multiple domains. As pointed out eg.\ by \newcite{Dredze08online}, this setting borrows much from DA, but also from multi-task learning \cite{Caruana97multitask}. If the intuitions behind MD may seem simple, the exact specifications of MDMT systems are rarely spelled out, which makes their evaluation difficult: for instance, should they perform well when the test data is distributed like the train data, when it is equally distributed across training domains or when the test distribution is unknown? should MDMT also be robust to new domains? How should they handle domain labeling errors?

@@

\section{Requirements of multi-domain MT \label{sec:requirements}}
In this section, we try to recap the main reasons for considering a multi-domain scenario and discuss their implications in terms of performance evaluation.
\fyTodo{Relation to theory, discussion of worst case scenario}

\subsection{MDMT: a tentative definition}
Informally, a multi-domain system should be able to accurately translate heterogeneous sources of data. This loose definition correspond to several practical situations:
\begin{itemize}
\item the more common situation is when both train and test documents (or sentences) are domain-tagged, where each domain corresponds to a specific source of data, and differ from the other sources in terms of textual genre, thematic content \cite{Zhang16topicinformed}, register, style \cite{Niu18multitask}, etc. It is usually assumed that the set of test domains is the same as the set of training domains -- when this is not the case, we would have zero-shot domain adaptation, similar to what is studied in multilingual MT \cite{Firat16multiway,Ha16towards,Johnson17google,Platanios18contextual}. This is the more general setting, already proposed in \cite{Dredze09multidomain}.\fyTodo{do we need a name here: such as supervised MDMT ? and mostly supervised MDMT for the second case ?}

@@

\item[[M2\hspace{-0.5em}]] statistical arguments. The training data available for each domain is very unevenly distributed, and domain trained or adapted on small datasets will yield large estimator variance. Building mix-domain systems is likely to reduce this variance, at the expense of a larger statistical bias \cite{Clark12onesystem}. Under this view, MDMT would be especially useful for domains with little training data. This is what is observed for multilingual MT from English: an increase for under-resource languages due to positive transfer, at the cost of a decrease for well-resourced languages \cite{Arivazhagan19massively}. A slightly distinct line of reasoning is that mixing domains has a positive regularization effect for all domains. By introducing variability in training, it prevents domain adaptation from overfitting the available adaptation data and could help improve generalization even for well-resourced domains.

@@

\paragraph{[P1-LAB]}\fyTodo{Decide naming scheme} A first expectation for MDMT is that they perform better that the obvious baseline which disregards domain labels. % Testing that MDMT actually benefit from domain tags is simple, notwithstanding the previous remarks above regarding evaluation procedures.

>>>>

Owing to the great variability of language data, this ideal situation is rarely met in practice, warranting the study of an alternative scenario, where the test distribution $\mathcal{D'}$ differs from $\mathcal{D}$. In this setting, \emph{domain adaptation} (DA) methods are required. DA has a long history in Machine Learning in general (eg.\ \cite{Shimodaira00improving,Ben-David09atheory}) and in NLP in particular (eg.\ \cite{Daume06domain,Blitzer07domain,Jiang07instance}), and various techniques exist to handle both the situations where a (small) training sample drawn from $\mathcal{D'}$ is available in training, or where only samples of source-side (or target-side) sentences are available (see \cite{Foster07mixture,Bertoldi09domain,Axelrod11domain} for early proposals of the statistical MT era, or \cite{Chu18asurvey} for a recent survey of DA in the context of Neural MT).
% In a nutshell, DA attempts to compensate for the distribution mismatch either by biasing the training sample to match ${\mathcal{D'}$ (eg.\ using resampling or instance-weighting strategies), or by adapting the

A seemingly related problem is \emph{multi-domain (MD) machine translation} \cite{Hassan17neural,Farajian17multidomain,Kobus17domaincontrol,Zeng18multidomain,Pham19generic} where one single system has to be trained and tested with data from multiple domains. As pointed out by \newcite{Dredze08online}, this setting borrows much from DA, but also from multi-task learning \cite{Caruana97multitask,Yang15unified}; note that it also share commonalties with ensemble learning \cite{XXX}. If the intuitions behind MDMT are simple, the exact specifications of MDMT systems are rarely spelled out, which makes their evaluation fragile: for instance, should they perform well when the test data is distributed like the train data, when it is equally distributed across training domains or when the test distribution is unknown? Should MDMT also be robust to new domains? How should they handle domain labeling errors? etc.

@@

\section{Requirements of multi-domain MT \label{sec:requirements}}
In this section, we try to recap the main reasons for considering a multi-domain scenario and discuss their implications in terms of performance evaluation.
\fyTodo{Relation to theory, discussion of worst case scenario}
\fyTodo{Insist on the work of Dredze}

\subsection{Formalizing multi-domain learning \label{ssec:formalization}}

We conventionally define a domain as a joint distribution $\mathcal{D}_d(x,y)$ over input-output pairs in $(x,y) \in \mathcal{X}\times{}\mathcal{Y}$. Assuming that the training data is sampled from a mixture of $D$ source domains $\mathcal{D}_s(x,y) = \sum_d \lambda_{s,d} \mathcal{D}_d(x,y)$, several cases are studied in the ML literature:

\begin{itemize}
\item the test distribution $\mathcal{D}_t$ is known 
  \begin{itemize}
  \item $\mathcal{D}_t = \mathcal{D}_{s}$ - this is standard ML, with mixture models
  \item $\mathcal{D}_t(x,y) = \sum_d \lambda_{t,d} \mathcal{D}_d(x,y) $ with $\lambda_s \neq \lambda_t$, a scenario deemed \emph{multiple source adaptation} (MSA) \cite{Mansour09domainadaptation}, with a different mixture of domains in training and testing
  \item $\mathcal{D}_t(x,y) \neq \sum_d \lambda_{t,d} \mathcal{D}_d(x,y) \forall \lambda$, a more complex case of MSA initially studied in \cite{Mansour09multiple}
  \item variants of cases 3) and 4) only assume knowledge of the marginal distribution $\mathcal{D}_t(x) = \sum_{y} = \mathcal{D}_t(x) $
  \item a slightly less favorable situation is when the test and train distributions are estimated from a finite sample, also studied in \cite{Mansour09multiple}.
  \item we recover standard supervised DA is when there is only one active domain in the training and test mixtures, and multiple-source single domain adaptation when the source is a mixture but the test domain is homogeneous. Unsupervised DA \cite{Daume06domain} corresponds to the case where only the marginal distribution of the test domain is available, while semi-supervised DA is the reverse situation.
  \end{itemize}
\item the test distribution is not known and training needs to be distributionally robust, ie.\ minimize the error against all possible distributions, with variants depending on whether it is or not assumed to be a mixture of existing domains, and whether it is perfectly or imperfectly known. This setting is also studied in eg.\ \cite{Mansour09multiple,Hoffman18algorithms}. This adversarial setting is also considered for training language models in \cite{Oren19distributionally}.
\item with respect to the labelling function $f$:
  \begin{itemize}
  \item $y=f(x)$ is deterministic, meaning that the joint distribution is entirely specified by the marginal and $f$; a more complex case is when $f$ also varies across domains;  
  \item the labelling function $P(y|x)$ is stochastic, but shared across all domains, which is the typical covariate-shift scenario; it can also vary across domains \cite{???}
  \end{itemize}
\end{itemize}

Multi-domain learning, as defined in \cite{Dredze08online} additionally makes the requirement that categorial domain tags are available in training and testing and that the learner should use them effectively \cite{Joshi12multidomain}. This is also the view that we adopt here, where we assume training inputs in $(\mathcal{X}\times{}\{1\dots{}D\})$, sampled from $\sum \lambda_{s,d}\mathcal{D}_{d} \times \{d\}$, and also consider various test scenarios, depending whether $\mathcal{D}_t$ is a fixed mixture of known domains, include new domain(s), in the supervised and unsupervised, and zero-shot settings.\fyTodo{explain zero shot}.

\subsection{MDMT: a tentative definition \label{ssec:definition}}
Informally, a multi-domain system should be able to accurately translate heterogeneous sources of data. This loose definition correspond to several practical situations:
\begin{itemize}
\item the more common situation is when both train and test documents (or sentences) are domain-tagged, where each domain corresponds to a specific source of data, and differ from the other sources in terms of textual genre, thematic content \cite{Zhang16topicinformed}, register, style \cite{Niu18multitask}, etc. It is usually assumed that the set of test domains is the same as the set of training domains -- when this is not the case, we would have zero-shot domain adaptation, similar to what is studied in multilingual MT \cite{Firat16multiway,Ha16towards,Johnson17google,Platanios18contextual}. This is the more general setting, already proposed in \cite{Dredze09multidomain}.\fyTodo{do we need a name here: such as supervised MDMT ? and mostly supervised MDMT for the second case ?}

@@

\item[[M2\hspace{-0.5em}]] statistical arguments. The training data available for each domain is very unevenly distributed, and domain trained or adapted on small datasets will yield large estimator variance. Building mix-domain systems is likely to reduce this variance, at the expense of a larger statistical bias \cite{Clark12onesystem}. Under this view, MDMT would be especially useful for domains with little training data. This is what is observed for multilingual MT from English: an increase for under-resource languages due to positive transfer, at the cost of a decrease for well-resourced languages \cite{Arivazhagan19massively}. A slightly distinct line of reasoning is that mixing domains has a positive regularization effect for all domains. By introducing variability in training, it prevents domain adaptation from overfitting the available adaptation data and could help improve generalization even for well-resourced domains. Following \cite{Joshi12multidomain}, one could even expect a benefit from ensembling several MT systems in decoding, irrespective of domain tags.

@@

\paragraph{[P1-LAB]}\fyTodo{Decide naming scheme} A first expectation for MDMT is that they perform better that the obvious baseline which would disregards domain labels, or re-assign them in a random fashion \cite{Joshi12multidomain}. % Testing that MDMT actually benefit from domain tags is simple, notwithstanding the previous remarks above regarding evaluation procedures.
